Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} Model System

### 2014-05-05 paper_enahpc14/Conclusion.tex
updates working copy

We  have   presented  a  methodology  for   comparing  performance  of
COSMO-ART,  a  regional  weather  forecast  model  augmented  for  the
interactions of  reactive gases and aerosol  particles.  The resulting
benchmarks illustrate  that the  best time-to-solution does  not imply
the best energy-to-solution: an Intel Sandybridge (2.6 GHz) system has
lower  time-to-solution but  higher energy-to-solution  than  an Intel
Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
strongly correlated.  On the other hand, the  usage of energy-friendly
MPI waiting  techniques analzyed with  our power-performance framework
over  a  series  of runs  of  \cosmoart  on  \tinto and  power  traces
demonstrate that it  is possible to reduce both  power dissipation and
energy-to-solution   while  maintaining   (or  even   decreasing)  the
time-to-solution. The resulting profiles indicate that simple changes,
such as making use of the  blocking MPI policy rather than the polling
is  possible to  slightly  reduce both  power  dissipation and  energy
consumption (by 2\,\%).

>>>>

We  have   presented  a  methodology  for   comparing  performance  of
COSMO-ART,  a  regional  weather  forecast  model  augmented  for  the
interactions of  reactive gases and aerosol  particles.  The resulting
benchmarks illustrate  that the  best time-to-solution does  not imply
the best energy-to-solution: an Intel Sandybridge (2.6 GHz) system has
lower  time-to-solution but  higher energy-to-solution  than  an Intel
Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
strongly correlated.  On the  other hand, the usage of energy-friendly
MPI waiting  techniques analysed with  our power-performance framework
over  a  series  of runs  of  \cosmoart  on  \tinto and  power  traces
demonstrate that it  is possible to reduce both  power dissipation and
energy-to-solution   while  maintaining   (or  even   decreasing)  the
time-to-solution. The resulting profiles indicate that simple changes,
such as making use of the  blocking MPI policy rather than the polling
is  possible to  slightly  reduce both  power  dissipation and  energy
consumption (by 2\,\%).

### 2014-05-05 paper_enahpc14/Introduction.tex
updates working copy

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
scale  modelling  system  is  coupled  with  the  operational  weather
forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed
and  used by  a consortium  of European  weather centers,  as  well as
utilized in a climate version by the wider research community.

The extended  atmospheric model \cosmoart  is com\-put\-ationally much
more demanding than \textsc{Cosmo}  since a large number of additional
tracers  and processes  have  to  be considered.   Thus  the model  is
currently severely limited in  terms of applicability and expensive in
terms  of energy  consumption.  Although  \textsc{Cosmo}  has recently
been  ported  to  GPUs  \cite{Gysi-2014, Lapillonne-2014}  within  the
framework  of the  High  Performance and  High Productivity  Computing
(HP2C)  Initiative \cite{HP2C}  to optimize  it for  computational and
energy efficiency,  significant investments in ART  are still required
to take it  to a similar level.  The efficiency  of \cosmoart is being
addressed in  the EU Exa2Green  project \cite{EXA2GREEN} to  deliver a
prototype code, which  provides an energy efficiency of  at least five
times  the baseline  value.  Such  an implementation  would  allow the
community to  investigate critical questions at  higher resolution and
over longer periods, at reduced cost to the environment.

This work is  organized as follows: in Section~\ref{sec:1}  we give an
overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly
introduce  \cosmoart and specify  its technical  setup related  to the
investigated   performance   and   energy  evaluation   methods.    In
Section~\ref{sec:3}  we  describe  HPC  platforms,  power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking  study.   Section~\ref{sec:4}  presents  performance  and
energy  requirements  of  the  baseline  on  these  architectures  and
highlights  areas  where  improvements   will  be  necessary  for  the
subsequent   baseline   refactoring.     Finally,   we   conclude   in
Section~\ref{concl} with  some implications for  the Exa2Green project
and give an outlook for future research.

>>>>

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
scale  modelling  system  is  coupled  with  the  operational  weather
forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed
and  used by  a consortium  of European  weather centers,  as  well as
utilised in a climate version by the wider research community.

The extended  atmospheric model \cosmoart  is com\-put\-ationally much
more demanding than \textsc{Cosmo}  since a large number of additional
tracers  and processes  have  to  be considered.   Thus  the model  is
currently severely limited in  terms of applicability and expensive in
terms  of energy  consumption.  Although  \textsc{Cosmo}  has recently
been  ported  to  GPUs  \cite{Gysi-2014, Lapillonne-2014}  within  the
framework  of the  High  Performance and  High Productivity  Computing
(HP2C)  Initiative \cite{HP2C}  to optimise  it for  computational and
energy efficiency,  significant investments in ART  are still required
to take it  to a similar level.  The efficiency  of \cosmoart is being
addressed in  the EU Exa2Green  project \cite{EXA2GREEN} to  deliver a
prototype code, which  provides an energy efficiency of  at least five
times  the baseline  value.  Such  an implementation  would  allow the
community to  investigate critical questions at  higher resolution and
over longer periods, at reduced cost to the environment.

This work is  organised as follows: in Section~\ref{sec:1}  we give an
overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly
introduce  \cosmoart and specify  its technical  setup related  to the
investigated   performance   and   energy  evaluation   methods.    In
Section~\ref{sec:3}  we  describe  HPC  platforms,  power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking  study.   Section~\ref{sec:4}  presents  performance  and
energy  requirements  of  the  baseline  on  these  architectures  and
highlights  areas  where  improvements   will  be  necessary  for  the
subsequent   baseline   refactoring.     Finally,   we   conclude   in
Section~\ref{concl} with  some implications for  the Exa2Green project
and give an outlook for future research.

### 2014-05-05 paper_enahpc14/Sec2:COSMO-ART.tex
updates working copy

Three-dimensional simulations are performed over large parts of Europe
and the Mediterranean~Sea for April 13$^{th}$ 2010, near the equinox, in order to approximately
have a half  day  of  sun  exposure and  therefore  ensure  a typical
activation of the chemistry cycle.  The calculation domain corresponds
to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
points with  a horizontal resolution of  $0.22\,^{\circ}$, i.e., 50~km
in  both directions  and 40~vertical  layers.  These  24-hour forecast
simulations  are not  preceded  by  a spin-up  phase  to build up  the
simulated gaseous and  aerosol concentrations.  This condition means 
that the model initialization needs a period of adjustment to reach its equilibrium  state and minimize  the effect of
initial conditions for gases and aerosols on the model predictions.

@@

The considered  \cosmoart model  system is configured  with a
semi-Lagrangian    horizontal   advection    sche\-me    with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\cite{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
PreProcessor solver (KPP) for  the solution of atmospheric chemistry
ordinary  differential equations \cite{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterizations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

>>>>

Three-dimensional simulations are performed over large parts of Europe
and the Mediterranean~Sea for  April 13$^{th}$ 2010, near the equinox,
in  order  to  approximately have  a  half  day  of sun  exposure  and
therefore  ensure a typical  activation of  the chemistry  cycle.  The
calculation  domain  corresponds to  the  CORDEX-EU-44  domain and  is
covered  by  a  grid  of  $222\times 216$  points  with  a  horizontal
resolution  of $0.22\,^{\circ}$,  i.e., 50~km  in both  directions and
40~vertical  layers.   These  24-hour  forecast  simulations  are  not
preceded  by a spin-up  phase to  build up  the simulated  gaseous and
aerosol   concentrations.   This  condition   means  that   the  model
initialization needs  a period of adjustment to  reach its equilibrium
state  and minimise  the effect  of initial  conditions for  gases and
aerosols on the model predictions.

@@

The   considered  \cosmoart   model  system   is  configured   with  a
semi-Lagrangian   horizontal    advection   sche\-me   with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\cite{COSMO-PartI-2011}.    It   also  makes   use   of  the   Kinetic
PreProcessor solver  (KPP) for  the solution of  atmospheric chemistry
ordinary  differential equations  \cite{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterisations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

### 2014-05-05 paper_enahpc14/paper.tex
Manuel's changes
WS: epsfig
updates working copy

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
\documentclass[twocolumn]{svjour3}
%
\smartqed
%
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[round, sort&compress, numbers]{natbib}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{units}
\usepackage{color}
\usepackage{colortbl}
\usepackage{xspace}
%
\DeclareRobustCommand\IPCClongname{}
\setcounter{secnumdepth}{5}
%
% please place your own definitions here and don't use \def but \newcommand{}{}
\newcommand{\pmlib}{\texttt{pmlib}\xspace}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Evaluating the Performance and Energy Efficiency of the
  COSMO-ART model system}
%\thanks{}
%\subtitle{}

@@

\begin{abstract}
  In this  paper we investigate  the energy footprint  and performance
  profiling of COSMO-ART  on various HPC platforms.  This  model is an
  extension of  the operational weather  forecast model of  the German
  weather  service   (DWD),  developed  for  the   evaluation  of  the
  interactions of reactive gases  and aerosol particles with the state
  of atmosphere at the  regional scale.  Different measurement devices
  and energy-aware techniques are  described to evaluate both time and
  energy  to  solution  of  the  considered application  and  to  gain
  detailed  insights  into power  and  performance requirements.   Our
  motivation  is to  improve  corresponding code  sections to  sustain
  performance  while minimising energy-to-solution.   This preliminary
  work  sets the  basis for  subsequent studies  to  tackle challenges
  related  to  energy  efficient  high performance  computing  in  the
  framework of the Exa2Green project \citep{EXA2GREEN}.

\keywords{High performance computing  \and Energy-aware computing \and
  Green computing  \and Numerical weather  prediction \and Atmospheric
  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
  Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
  Time-to-solution \and Energy-to-solution \and Xeon processor}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

\section{Related work}
\label{sec:1}
\input{Sec1:Related.tex}

\section{The COSMO-ART model system}
\label{sec:2}
\input{Sec2:COSMO-ART.tex}

\section{Environment setup}
\label{sec:3}
\input{Sec3:Environment-setup.tex}

\section{Experimental results}
\label{sec:4}
\input{Sec4:Experimental-results.tex}

\section{Conclusion}
\label{concl}
\input{Conclusion.tex}

@@

\begin{acknowledgements}
The research  leading to these  results is supported by  the Exa2Green
project  co-financed by  the European  Commission under  7th Framework
Programme Future and Emerging Technologies (FET) Proactive Initiative:
Minimising Energy  Consumption of Computing to the  Limit (MINECC). We
also gratefully acknowledge the High Performance and High Productivity
Computing Initiative  \citep{HP2C} for results that  will be leveraged
in subsequent code refactoring.
\end{acknowledgements}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
\documentclass[twocolumn]{svjour3}
%
\smartqed
%
\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{colortbl}
\usepackage{url}
\usepackage{xspace}
%
\DeclareRobustCommand\IPCClongname{}
%
% please place your own definitions here and don't use \def but \newcommand{}{}
\newcommand{\pmlib}{\texttt{pmlib}\xspace}
\newcommand{\monch}{\textsc{Monch}\xspace}
\newcommand{\pilat}{\textsc{Pilatus}\xspace}
\newcommand{\tinto}{\textsc{Tintorrum}\xspace}
\newcommand{\cosmoart}{\textsc{Cosmo-Art}\xspace}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Evaluating  the  Performance   and  Energy  Efficiency  of  the
  \textsc{COSMO-ART} Model System}
%\thanks{}
%\subtitle{}

@@

\begin{abstract}
  In this  paper we investigate  the energy footprint  and performance
  profiling of \cosmoart  on various HPC platforms.  This  model is an
  extension of  the operational weather  forecast model of  the German
  weather  service   (DWD),  developed  for  the   evaluation  of  the
  interactions of reactive gases  and aerosol particles with the state
  of atmosphere at the  regional scale.  Different measurement devices
  and energy-aware techniques are  described to evaluate both time and
  energy  to  solution  of  the  considered application  and  to  gain
  detailed  insights  into power  and  performance requirements.   Our
  motivation  is to  improve  corresponding code  sections to  sustain
  performance  while minimising energy-to-solution.   This preliminary
  work  sets the  basis for  subsequent studies  to  tackle challenges
  related  to  energy  efficient  high performance  computing  in  the
  framework of the Exa2Green project \cite{EXA2GREEN}.

\keywords{High performance computing  \and Energy-aware computing
  %Green computing  
  \and Numerical weather  prediction \and Atmospheric
  chemistry  \and  Aerosol  modelling \and
  Benchmark  analysis  \and 
  \cosmoart  coupled  model \and  
  Power-performance profiling and tracing tools  \and
  Energy-saving techniques}
% \PACS{PACS code1 \and PACS code2 \and more} 
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

\section{Related Work}
\label{sec:1}
\input{Sec1:Related.tex}

\section{The \textsc{COSMO-ART} Model System}
\label{sec:2}
\input{Sec2:COSMO-ART.tex}

\section{Environment Setup}
\label{sec:3}
\input{Sec3:Environment-setup.tex}

\section{Experimental Results}
\label{sec:4}
\input{Sec4:Experimental-results.tex}

\section{Concluding Remarks}
\label{concl}
\input{Conclusion.tex}

@@

\begin{acknowledgements}
The research  leading to these  results is supported by  the Exa2Green
project  co-financed by  the European  Commission under  7th Framework
Programme Future and Emerging Technologies (FET) Proactive Initiative:
Minimising Energy  Consumption of Computing to the  Limit (MINECC). We
also gratefully acknowledge the High Performance and High Productivity
Computing Initiative \cite{HP2C} for results that will be leveraged in
subsequent code refactoring.
\end{acknowledgements}
